---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height = 5,
  fig_width = 8,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ialiquor

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![GitHub commit](https://img.shields.io/github/last-commit/nikdata/ialiquor)](https://github.com/nikdata/ialiquor/commit/main)
<!-- badges: end -->

The **{ialiquor}** package provides a summary of the monthly liquor sales by county and by liquor type in the State of Iowa. This dataset comes from the Iowa Data Portal and is limited to beverages classified as [Class E](https://abd.iowa.gov/license-classifications). Class E beverages (as according to Iowa) are:

> For grocery, liquor and convenience stores, etc. Allows for the sale of alcoholic liquor for off-premises consumption in original unopened containers. No sales by the drink. Sunday sales are included. Also allows wholesale sales to on-premises Class A, B, C and D liquor licensees but must have a TTB Federal Wholesale Basic Permit.

In other words, this dataset is limited to retailers that are essentially selling "hard" liquor. This does not include data for beer sales (which is a Class C license) or wine (which is a Class B license).

In Iowa, the state will purchase alcohol from vendors and then resell the alcohol to retailers within the state. As such, there is a 'cost' that the state pays to the vendor and then there is 'revenue' to the state based on the price sold to the retailer. It is important to keep in mind that these revenue values are not indicative of the actual sale price to the end customer.

**WIP**

## Installation

This package is not available on CRAN, but can be installed from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("nikdata/ialiquor", ref = "main")
```

## Usage

The {ialiquor} package provides a monthly summary (from January 2015 onward) of the alcohol (class E) sales in the State of Iowa. Several attributes are available in the dataset such as county, county population, the type (derived by me) of alcohol, the category of the alcohol (as defined by the State of Iowa). Furthermore, numerical summaries are available for cost, revenue, bottles sold, and volume.

### Preview of Liquor Sales

```{r}
library(ialiquor)
library(dplyr)
data('liquor_sales')

dplyr::glimpse(liquor_sales)

```


### Potential Use Cases

- Revenue by type
- Total revenue per bottle to state by year
- Total volume sold per person by year