---
title: "Overview of Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The {ialiquor} package contains two datasets that summarize the monthly Class E liquor sales (along with other attributes) in the State of Iowa.

The `county_sales` dataset contains the following fields:

- `year_mon` - The Year and month in YYYY-MM-DD format
- `county` - Name of county
- `county number` - County number (there are 99 counties in Iowa)
- `state_cost` - Amount paid for the liquor by the State of Iowa in US$ (not normalized)
- `bottles_sold` - Number of bottles sold (regardless of size of bottle or volume)
- `retail_revenue` - Amount paid by retail vendors for the alcohol in US$ (not normalized). This field tallies the total retail revenue for all retailers within the respective county.
- `liters_sold` - Volume of liquor sold (in liters)

The `category_sales` dataset contains the following fields:

- `year_mon` - The Year and month in YYYY-MM-DD format
- `type` - a high level categorization of the type of liquor (e.g., vodka, whiskey, rum, etc.)
- `category` - a second level categorization of the type of liquor (e.g., flavored vodka, scotch, bourbon, spiced rum, etc.)
- `state_cost` - Amount paid for the liquor by the State of Iowa in US$ (not normalized)
- `bottles_sold` - Number of bottles sold (regardless of size of bottle or volume)
- `retail_revenue` - Amount paid by retail vendors for the alcohol in US$ (not normalized). This field tallies the total retail revenue for all retailers within the respective county.
- `liters_sold` - Volume of liquor sold (in liters)

**Data Source**: Iowa Data Portal [API](https://data.iowa.gov/resource/m3tr-qhgy.csv)


```{r}
library(ialiquor)

str(county_sales)

head(county_sales)

```

```{r}
library(ialiquor)

str(category_sales)

head(category_sales)
```

